---
title: "Stepwise feature selection for regression"
teaching: 0
exercises: 0
questions:
- "Why would we want to find a subset of features
  that are associated with an outcome?"
- "How can we iteratively find a good subset of our features
  variables to use for regression?"
- "What are some risks and downsides of iterative feature
  selection?"
objectives:
- "Understand multiple regression in a biomedical context."
- "Understand how to fit a stepwise regression model."
keypoints:
- "Sets of features can be more predictive and provide
  a better explanation than a single feature alone."
- "Stepwise regression allows us to find a set of features that
  are associated with an outcome (eg, age)."
- "Stepwise regression will tend to retain only one
  feature out of many that are correlated."
- "Stepwise regression is not very efficient."
math: yes
---


```{r, include=FALSE}
library("here")
source(here("bin/chunk-options.R"))
knitr_fig_path("02b-")
```


In the previous 

Another way of modelling these data is to model age as 

$$
    y_j = \beta_0 + \beta_1 X_1 + \dots \beta_p X_p + \epsilon_j
$$


```{r}
suppressPackageStartupMessages({
    library("glmnet")
    library("limma")
    library("minfi")
    library("here")
    library("broom")
})

if (!file.exists(here("data/methylation.rds"))) {
    source(here("data/methylation.R"))
}
norm <- readRDS(here("data/methylation.rds"))

lim <- norm
y <- lim$Age
X <- getM(lim)
```

However when the number of predictors is greater than the number of samples
(basically always true in genetics) it isn't possible to include everything!

There are some techniques that you can use to find a set of predictors!

- screening (correlation etc): bad, don't do
- screening (variance): not necessarily bad if the screening variable is sensible
- forward/reverse/best subset selection

```{r}
if (!file.exists(here("data/synthetic.rds"))) {
    source(here("data/synthetic.R"))
}
synthetic <- readRDS(here("data/synthetic.rds"))
```




```{r}
X <- assay(synthetic)
y <- synthetic$age
beta <- rowData(synthetic)$true_beta
names(beta) <- rownames(synthetic)
## challenge 3: fit y on x univariate
## compare with true betas
cc <- sapply(seq_len(nrow(X)), function(i) {
    coef(lm(y ~ X[i, ]))[[2]]
})
plot(cc, beta, pch = 19, cex = 0.5)
abline(0, 1)
abline(v = 0, lty = "dashed", col = "firebrick")
abline(h = 0, lty = "dashed", col = "firebrick")
```




> ## Exercise
> Perform forward subset selection on the methylation data.
> 
> 
> > ## Solution
> > 
> > 
> {: .solution}
{: .challenge}



```{r}
## challenge 4: forward selection
## compare with true betas
xy <- as.data.frame(cbind(t(X), y = y))
int <- lm(y ~ 1, data = xy)
all <- lm(y ~ . + 0, data = xy)
forward <- step(
    int,
    scope = list(upper = formula(all), lower = formula(int)),
    direction = "forward",
    trace = 0
)
forward$anova
plot(coef(forward), beta[names(coef(forward))])
abline(0, 1)
abline(v = 0, lty = "dashed", col = "firebrick")
abline(h = 0, lty = "dashed", col = "firebrick")
```

```{r}
## note about backward/both, not a challenge
all <- lm(y ~ . + 0, data = xy)
backward <- step(
    all,
    scope = formula(all),
    direction = "backward",
    trace = 0
)
backward$anova
plot(coef(backward), beta[names(coef(backward))])
abline(0, 1)
abline(v = 0, lty = "dashed", col = "firebrick")
abline(h = 0, lty = "dashed", col = "firebrick")
```


{% include links.md %}
