---
title: "Hierarchical clustering"
teaching: 0
exercises: 0
questions:
- "What does clustering mean?"
- "Why would we want to find clusters in data?"
- "How can we cluster data with a model?"
- "How can we cluster data without a model?"
- "How can we check if clusters are robust?"
objectives:
- "Understand and perform clustering with K-means, mixture models, and hierarchical clustering."
- "Assess clustering performance with silhouette score and bootstrapping/consensus clustering."
keypoints:
- "KP1"
math: yes
---



# Hierarchical clustering

Recall the heatmap I showed in regression lesson.
If we do this without hierarchical clustering, it's very noisy.
```{r, fig.cap="Cap", fig.alt="Alt"}
library("minfi")
library("here")
library("ComplexHeatmap")

methylation <- readRDS(here("data/methylation.rds"))

age <- methylation$Age
methyl_mat <- t(assay(methylation))
small <- methyl_mat[, 1:500]
cor_mat <- cor(small)
col <- circlize::colorRamp2(
    breaks = seq(-1, 1, length.out = 9),
    colors = rev(RColorBrewer::brewer.pal(9, "RdYlBu"))
)
Heatmap(cor_mat,
    column_title = "Feature-feature correlation in methylation data",
    name = "Pearson correlation",
    col = col,
    cluster_rows = FALSE, cluster_columns = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE,
    show_row_names = FALSE, show_column_names = FALSE
)
```



With clustering you can see some nice groupings.

```{r, fig.cap="Cap", fig.alt="Alt"}
Heatmap(cor_mat,
    column_title = "Feature-feature correlation in methylation data",
    name = "Pearson correlation",
    col = col,
    row_dend_width = unit(0.2, "npc"),
    column_dend_height = unit(0.2, "npc"),
    show_row_names = FALSE, show_column_names = FALSE
)
```

# Distance functions

# Linkage methods

Complete linkage (the default) uses the maximum distance between clusters as
the distance when merging them.

```{r, fig.cap="Cap", fig.alt="Alt"}
distmat <- dist(cor_mat)
clust <- hclust(distmat, method = "complete")
plot(clust)
```

Ward’s method says that the distance between two clusters, A and B, is how much
the sum of squares will increase when we merge them.

Details: https://jbhender.github.io/Stats506/F18/GP/Group10.html
